---
title: "Import data"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(stringr)
library(mxbstuff)

```


## Data for core samples

The raw data give values of carbon and nitrogen for analytical replicates sampled from each soil core. For most cores there were 3 replicates, but some had 5 and two had 19 and 22 respectively (intended to detect for analytical batch effects).

The data set for sample values also includes core-level measurements (top soil depth and core weight) together with site-level variables (fire history etc).

Standard samples are also included, with a text string (`standard` column) indicating the composition of the standard. The string is empty for field samples.

```{r}

dat <- read.csv(mdPath("data_raw/Eden_SoilData.csv"))

# Fix column names 
colnames(dat) <- colnames(dat) %>%
  tolower() %>%
  str_replace_all("_|\\.", "")

# Add data row number (for back reference). 
# Add a flag for field samples.
dat <- dat %>% 
  mutate(datarow = row_number(),
         sample = str_trim(standard) == "")

```


Separate out data for standards. Note standards were assigned a core number which we retain here just in case.

```{r}

STANDARDS <- dat %>%
  filter(!sample) %>%

  select(datarow, stickerno, coreno, standard, totalcarbon, totalnitrogen) %>%
  
  mutate(standard = factor(str_trim(standard))) %>%
  
  mutate(coreno = droplevels(coreno))

glimpse(STANDARDS)

```


Separate out data for field cores.

```{r}

SAMPLES <- dat %>%
  filter(sample) %>%

  select(datarow, plot, stickerno, coreno, treeopen, 
         coredepth, soildepth, topsoildepth, soilcolour, weightg, bulkdengcm3,
         totalcarbon, totalnitrogen) %>%

  mutate(coreno = droplevels(coreno),
         treeopen = droplevels(treeopen)) %>%

  # Set the few blank values for soil colour to NA
  mutate(soilcolour = as.character(soilcolour),
         soilcolour = factor(soilcolour, exclude = "")) %>%

  # Fix bulk density column which ends up as a factor because
  # of "DIV0"" flags for standards in the input data
  mutate(bulkdengcm3 = as.numeric( as.character(bulkdengcm3) ) )

glimpse(SAMPLES)

```


Separate out site-level data.

```{r}

SITES <- dat %>%
  filter(sample) %>%
  select(plot, treatment, tsf, firessince1986) %>%
  distinct() %>%
  mutate(treatment = droplevels(treatment),
         
         harvest = factor(str_sub(treatment, 1, 1), 
                         levels = c("U", "L"), labels = c("unharvested", "harvested")),
         
         firetreat = factor(str_sub(treatment, 2, 2),
                            levels = c("N", "R", "F"), labels = c("none", "regular", "frequent"))
  )

# Check that we have only one row for each site 
# (ie. there were no different values for site level variables)
x <- SITES %>% 
  group_by(plot) %>%
  summarize(N = n()) %>%
  filter(N > 1)

if (nrow(x) > 0) stop("Inconsistent data for sites")

# Fix numeric vars that have ended up as factors
SITES <- SITES %>%
  mutate(tsf = as.numeric( as.character(tsf) ), 
         firessince1986 = as.numeric( as.character(firessince1986) ))

glimpse(SITES)

```


### Check for missing values in field samples

```{r}

ii <- rowSums(is.na(SAMPLES)) > 0
x <- SAMPLES[ii, ] %>%
  select(datarow, everything())

knitr::kable(x)

```


Calculate sample means for carbon and nitrogen. 

**TODO: modify this to take into account the different numbers of replicates for different cores.**

```{r}

SAMPLE.MEANS <- SAMPLES %>%
  select(-datarow, -stickerno) %>%
  
  group_by(coreno) %>%
  
  mutate(nreps = n(),
         totalcarbon = mean(totalcarbon),
         totalnitrogen = mean(totalnitrogen)) %>%
  
  ungroup() %>%

  distinct(plot, coreno, .keep_all = TRUE) %>%
  
  select(plot, coreno, everything())

glimpse(SAMPLE.MEANS)

```


From the sample means, calculate tonnes per hectare values for carbon and nitrogen and C/N ratio based on these values.

```{r}

SAMPLE.MEANS <- SAMPLE.MEANS %>%
  
  mutate(carbontha = coredepth * bulkdengcm3 * totalcarbon,
         nitrogentha = coredepth * bulkdengcm3 * totalnitrogen) %>%
  
  mutate(cnratio = carbontha / nitrogentha)

```



## Additional data for sites

Data for site locations and topography were provided in an Excel file.

```{r}

x <- readxl::read_excel(mdPath("data_raw/EBSA_Site_data.xlsx")) %>%
  select(plot, easting, northing,
         aspect = bsaaspect2,
         slope = bsaslope25,
         elevation = bsadem25,
         wetness = bsawet25,
         solar = netrad)

SITES <- left_join(SITES, x, by = "plot")

# Add a 'northness' variable
SITES <- SITES %>%
  mutate(northness = sin(0.5 * pi * aspect / 180))

```

Fire data for plots.

```{r}

# burnp is percentage of sample points (usually n=10) recorded as burnt.
# Trent required 2 or points burnt for the plot be treated as burnt.

FIRES <- read.csv( mdPath("data_raw/fire_history_2013.csv"), stringsAsFactors = FALSE) %>%
  mutate(plot = as.integer(str_extract(plot_txt, "\\d+$"))) %>%
  select(plot, year, burnp) %>%
  mutate(burnt = as.integer(burnp >= 20))

```


Save data to file.

```{r}

save(STANDARDS, file = mdPath("data/standards.RData"))
save(SAMPLES, file = mdPath("data/samples.RData"))
save(SAMPLE.MEANS, file = mdPath("data/sample_means.RData"))
save(SITES, file = mdPath("data/sites.RData"))
save(FIRES, file = mdPath("data/fires.RData"))

```


